
\documentclass{article}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{amssymb}

%TCIDATA{OutputFilter=LATEX.DLL}
%TCIDATA{Version=4.00.0.2312}
%TCIDATA{Created=Thursday, February 05, 2009 18:41:07}
%TCIDATA{LastRevised=Sunday, February 08, 2009 23:33:51}
%TCIDATA{<META NAME="GraphicsSave" CONTENT="32">}
%TCIDATA{<META NAME="DocumentShell" CONTENT="Standard LaTeX\Blank - Standard LaTeX Article">}
%TCIDATA{CSTFile=40 LaTeX article.cst}

\newtheorem{theorem}{Theorem}
\newtheorem{acknowledgement}[theorem]{Acknowledgement}
\newtheorem{algorithm}[theorem]{Algorithm}
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem{case}[theorem]{Case}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{conclusion}[theorem]{Conclusion}
\newtheorem{condition}[theorem]{Condition}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{criterion}[theorem]{Criterion}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{problem}[theorem]{Problem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{solution}[theorem]{Solution}
\newtheorem{summary}[theorem]{Summary}
\newenvironment{proof}[1][Proof]{\noindent\textbf{#1.} }{\ \rule{0.5em}{0.5em}}
\input{tcilatex}

\begin{document}


\ref{FigYC1M}

\FRAME{ftbpFU}{254.8125pt}{398.9375pt}{0pt}{\Qcb{Yield curve on Euribor1M.
Brown line: zero curve, plotted with zero rates $Z_{f}^{I_{1M}}\left(
t_{0},t_{0}+t,\text{\textit{act/365}}\right) $, $t$ daily sampled and
settlement date $t_{0}=$ Jan. 27th, 2009. Blu line: forward curve $%
\complement _{f}^{I_{1M}}$, plotted with 1M-tenor forward rates $F\left(
t_{0};t,t+1M,\text{\textit{act/360}}\right) $. Upper panel: short term
structure up to 3 years; middel panel: medium term structure up to 10 years;
lower panel: long term structure up to 60 years.}}{\Qlb{FigYC1M}}{Figure}{%
\special{language "Scientific Word";type "GRAPHIC";maintain-aspect-ratio
TRUE;display "USEDEF";valid_file "T";width 254.8125pt;height
398.9375pt;depth 0pt;original-width 339.5pt;original-height 533pt;cropleft
"0";croptop "1";cropright "1";cropbottom "0";tempfilename
'FigYC1M.wmf';tempfile-properties "XNPR";}}

\FRAME{ftbpFU}{254.8125pt}{398.9375pt}{0pt}{\Qcb{Yield curve on Euribor3M.
As in fig. \protect\ref{FigYC1M}. Cyan diamonds: quoted 3M FRAs. Green
triangles: quoted 3M\ Futures.}}{\Qlb{FigYC3M}}{Figure}{\special{language
"Scientific Word";type "GRAPHIC";maintain-aspect-ratio TRUE;display
"USEDEF";valid_file "T";width 254.8125pt;height 398.9375pt;depth
0pt;original-width 339.5pt;original-height 533pt;cropleft "0";croptop
"1";cropright "1";cropbottom "0";tempfilename
'FigYC3M.wmf';tempfile-properties "XNPR";}}

\FRAME{ftbpFU}{254.8125pt}{398.9375pt}{0pt}{\Qcb{Yield curve on Euribor6M.
As in fig. \protect\ref{FigYC1M}. Cyan diamonds: quoted 6M FRAs. Green
triangles: quoted 6M\ Futures.}}{\Qlb{FigYC6M}}{Figure}{\special{language
"Scientific Word";type "GRAPHIC";maintain-aspect-ratio TRUE;display
"USEDEF";valid_file "T";width 254.8125pt;height 398.9375pt;depth
0pt;original-width 339.5pt;original-height 533pt;cropleft "0";croptop
"1";cropright "1";cropbottom "0";tempfilename
'FigYC6M.wmf';tempfile-properties "XNPR";}}

\FRAME{ftbpFU}{254.8125pt}{398.9375pt}{0pt}{\Qcb{Yield curve on Euribor12M.
As in fig. \protect\ref{FigYC1M}. Cyan diamonds: quoted 12M FRAs.}}{\Qlb{%
FigYC12M}}{Figure}{\special{language "Scientific Word";type
"GRAPHIC";maintain-aspect-ratio TRUE;display "USEDEF";valid_file "T";width
254.8125pt;height 398.9375pt;depth 0pt;original-width
339.5pt;original-height 533pt;cropleft "0";croptop "1";cropright
"1";cropbottom "0";tempfilename 'FigYC12M.wmf';tempfile-properties "XNPR";}}

\end{document}
